\chapter{Background}
\label{ch:background}

\section{Proofs, verification \& formalization}
\label{sec:proofs_formalization}

When solving mathematical problems, we often use proofs
to either \textbf{justify} a claim or to \textbf{explain} why the claim is true.
We can distinguish between two types of proofs; \emph{informal} and \emph{formal} proofs.

An informal proof is often written in a natural language, and the proof is adequate if most
readers are convinced by the proof \cite[ch.~2]{bpierce}. Such proofs rely heavily on the reader's intuition
and often omit logical steps to make them easier to understand for humans \cite[p.~1371]{ams:formal-proof}.
As these proofs grow larger and more complex, they become harder for humans to follow,
which can ultimately lead to errors in the proofs' logic. This might cause the whole proof
to be incorrect \cite{rkhamsi}, and even the claim "justified" by the proof might be wrong.

A formal proof is written in a formal language, and can be compared to a computer program
written in a programming language. Writing a formal proof is more difficult than writing an informal proof.
Formal proofs include every logical step, and nothing is left for the reader (often a computer) to assume.
This can make formal proofs extremely verbose, but the amount of logical errors is reduced \cite[p.~1371]{ams:formal-proof}.
The only possible errors in formalized proofs are false assumptions and/or flawed verification software.
When we talk about \emph{formalization}, we mean the process of translating an informal proof into a formal proof.

\section{Type theory}
\label{sec:type_theory}

Type theory is a foundation of mathematics \cite{typetheorystan}
in which every mathematical object is viewed as being a member of a certain type.
We can say that the object $a$ is a member of the type $A$,
or, that $a$ has type $A$, denoted by $a : A$.
For any type $A$, we must know how to construct an element $a$ of that type \cite[p.~76]{lof1}.
A simple example would be the type of natural numbers, $\mathbb{N}$.
Any object $n : \mathbb{N}$ would be a natural number,
which we can easily provide by iterating a successor function on the constant $O$.
Type theory gives us a framework to work with and manipulate objects and types
by providing us with rules of inference for objects and their types.
There are several different type theories, where each one has their own rules.
As mathematical objects can be a lot more complex than just natural numbers,
the types these objects belong to follow in complexity.

A related concept in computer programming is the notion of a \emph{data type}.
Akin to type theory, these are often used to represent some object or structure,
e.g. the data type \texttt{bool} is usually used to represent a boolean value.

\subsection{Propositions as types}
\label{ssec:propositions_as_types}

The concept of proposition as types sees any proposition $P$ as representing a type,
and every proof $p$ of $P$ is an element of the type $P$, or $p : P$.
Thus, if we can construct an element of the type $P$, we can prove $P$,
and vice versa; if we can prove $P$, we can construct an element of the type of $P$.
For example, to prove a proposition $P$ which states "all natural numbers are the sum of four squares",
we must construct a value of the type $P$ that shows that this is true for all natural numbers.
Such a value is a function that for any input $n$ returns a proof that $n$ is the sum of four squares,
that is, return four numbers $a, b, c, d$ and a proof that $n = a^2 + b^2 + c^2 + d^2$.
Proofs are mathematical objects; thus a proposition can be viewed as having the type of all its proofs (if any!).
We can use this correspondence to model a proof as a typed computer program.
The power of this concept comes from the fact that we can use a type checker to verify that
our program is typed correctly, and thus that the corresponding proof is valid.
Often, the proof can be used to compute something, i.e., the numbers $a, b, c, d$ mentioned above
\cite{pwadler} \cite{lof2}.

\subsection{Dependent types}
\label{ssec:dependent_types}

Dependent types allow us to define more rigorously types that depend on values.
We again look at the proposition $P$ from above.
It states that for all natural numbers $n$,
there exists four other natural numbers $a, b, c, d$,
such that $P$ holds.
Or, in mathematical notation:
$$
\forall n \, \exists a \, \exists b \, \exists c \, \exists d \, (n = a^2 + b^2 + c^2 + d^2) \;\;\;
\text{where} \; n, a, b, c, d \in \mathbb{N}
$$
The proof of this proposition is, as stated previously,
a function that takes an argument $n$ and returns a proof, or,
evidence in the form of values for $a, b, c, d$ that satisfy $P$.
Since propositions are just types, any function $f : A \rightarrow B$
will produce a proof of $B$ given a proof of $A$.

To represent the type of an universal quantified proposition,
we use a $\Pi$-type, also called a dependent function type.
Given a function $B$ that for any $a : A$ maps to an
element of the type $B(a)$, we denote $\prod_{a : A} B(a)$.
This type is the type of all functions that for any element
$a : A$ maps to an element $b : B(a)$ \cite[p.~78] {lof1}.
In the case for $P$, the type $A$ would be $\mathbb{N}$,
and the element $a$ would be a natural number, corresponding to the variable $n$ in $P$.
A proof of $\prod_{n : \mathbb{N}} \, (\exists a \, \exists b \, \exists c \, \exists d \, (n = a^2 + b^2 + c^2 + d^2)$
is a function that for any $n \in \mathbb{N}$
gives us a proof of $\exists a \, \exists b \, \exists c \, \exists d \, (n = a^2 + b^2 + c^2 + d^2)$.
Thus, the type of the proposition varies with $n$,
modeling universal quantification.

To represent the type of an existentially quantified proposition,
we use a $\Sigma$-type, also called a dependent pair type.
Given a function $B$ that for any $a : A$ maps to an
element of the type $B(a)$, we denote $\sum_{a : A} B(a)$.
This type is the type of all pairs $(x, y)$ where
$x : A$ and $y : B(a)$ \cite[p.~79]{lof1}.
Again, in the case for $P$, $A$ would be $\mathbb{N}$,
and the element $a$ would be a natural number, corresponding to the variable $a$ in $P$.
A proof of $\prod_{n : \mathbb{N}} \, \sum_{a : \mathbb{N}} \, (\exists b \, \exists c \, \exists d \, (n = a^2 + b^2 + c^2 + d^2)$
is a pair $(x, y)$, where $x$ is evidence of $P$
and $y$ is a proof of $P$.
Hence, any evidence of $P$ is provided with its proof,
modeling existential quantification.
The other existential quantifiers are likewise replaced by $\Sigma$'s:
$$
\prod_{n : \mathbb{N}} \, \sum_{a : \mathbb{N}} \, \sum_{b : \mathbb{N}} \, \sum_{c : \mathbb{N}} \, \sum_{d : \mathbb{N}} \, (n = a^2 + b^2 + c^2 + d^2)
$$
The proposition $n = a^2 + b^2 + c^2 + d^2$ is a type dependent on the five values $n, a, b, c, d$.

As we have just shown by representing universal and existential quantification,
the concept of dependent types is important for propositions as types
by providing us with more expressivity.
We can also look at a more practical example of dependent types,
using the Coq proof assistant, which has support for dependently typed programming.
We consider the type \lstinline{vector}, which is a type polymorphic (the elements of the list can be of any same type)
finite list with a fixed length, which is the value on which the type depends, a natural number.

\begin{minipage}{\linewidth}
\begin{lstlisting}[language=Coq, label={lst:dep_type_vec}, caption={\lstinline{vector} in Coq, using dependent types}]
(* brackets around 'A : Type' indicate that its type should be inferred *)
Inductive vector {A : Type} : nat -> Type :=
  | Vnil : vector 0
  (* 'forall' is an instance of a dependent function type *)
  | Vcons : forall (h : A) (n : nat), vector n -> vector (S n).
\end{lstlisting}
\end{minipage}

This definition gives us a type with two constructors:

\begin{itemize}
    \item
        \lstinline{Vnil} has the type of \lstinline{vector 0}, and represents the empty vector.
    \item
        \lstinline{Vcons} has type of \lstinline{forall (h : A) (n : nat) vector (S n)}, where \lstinline{h}
                            is the head of the vector and \lstinline{n} is the length of the vector,
                            where both are given to the constructor as arguments.
\end{itemize}

In this scenario, the length of an object of the type \lstinline{vector n} is fixed by the argument \lstinline{n : nat}.
Any definition of a \lstinline{vector} must adhere to this term, and is checked at compile time.
An example of a valid and invalid definition is:

\begin{minipage}{\linewidth}
\begin{lstlisting}[language=Coq, label={lst:dep_type_vec_ex}, caption={Examples of vectors in Coq}]
(* valid definition; (S 0) equal to 1  *)
Definition vec_valid : vector 2 :=
    Vcons "b" 1 (Vcons "a" 0 (Vnil)).
(* invalid definition; (S 0) not equal to 2 *)
Fail Definition vec_invalid : vector 2 :=
    Vcons "b" 2 (Vcons "a" 0 (Vnil)).
\end{lstlisting}
\end{minipage}

\section{Proof assistants}
\label{sec:proof_assistants}

Propositions as types allow us to bridge the gap between logic and computing,
while dependent types allow us to define more rigorously types which depend on values,
thus enhancing the expressivity of simple types.
The former is a crucial aspect of \emph{proof assistants}, while the latter gives
us more expressive power when constructing proofs using a proof assistant,
as we saw when modeling universal and existential quantifiers using types.
An example of the expressive power of dependent types is the fact that we can define
predicates that depend on the value of a term, e.g. a predicate that checks if a number is even.
The purpose of a proof assistant is to get computer support to construct and verify a formal proof mechanically.

\subsection{Coq}
\label{ssec:coq}

Coq is based on the higher-order type theory \emph{Calculus of Inductive Constructions} (CIC), and
functions as both a proof assistant and a dependently typed functional programming language.
Coq also allow us to extract certified programs from the proofs of their specification to
the programming languages OCaml and Haskell \cite{cintro}.
Coq implements a specification language called \emph{Gallina},
which allows us to define logical objects within Coq.
These objects are each assigned a type to ensure they are correct,
and the rules used to assign these types come from CIC \cite{cic}.

This is an example of the syntax of Gallina:

\begin{minipage}{\linewidth}
\begin{lstlisting}[language=Coq, label={lst:gallina_ex}, caption={Example of Gallina syntax}]
Inductive nat : Type :=
  | O
  | S : nat -> nat.

Inductive le (n : nat) : nat -> Prop :=
  | le_n : n <= n
  | le_S : forall m : nat, n <= m -> n <= S m.

Definition lt_n_S_n :=
  (fun n : nat => le_n (S n)) : forall n : nat, n < S n.
\end{lstlisting}
\end{minipage}

Looking at the final definition in the example, we can see the
concept of propositions as types in action.
\lstinline{lt_n_S_n} defines a function which takes a natural number \lstinline{n}
as input, and returns a value of the type \lstinline{forall n : nat, n < S n},
denoted by the colon before the type itself.
The return value is therefore a proof of \lstinline{forall n : nat, n < S n}, and since
the definition has been type-checked by Coq, we know that this proof is valid!
In this case, the function is
\lstinline{fun : nat => le_s (S n)}, where \lstinline{le_n} is a constructor
of the type \lstinline{forall n : nat, n <= n}. By applying this constructor
to \lstinline{S n}, we get a value of the type (and a proof) of
\lstinline{forall n : nat, S n <= S n}.
By Coq's definition of \lstinline{<},
our initial theorem can be rewritten as \lstinline{forall n : nat, S n <= S n}.
This matches the type of our function, and the proof is complete.

Proving theorems like this is not really intuitive for a human prover,
and that is why Coq gives us the \emph{Ltac} meta-language for writing proofs.
Ltac provides us with tactics, which are a kind of shorthand syntax for defining Gallina terms \cite{cltac}.
Using Ltac, we can rewrite the proof from \Cref{lst:gallina_ex} as such:

\begin{minipage}{\linewidth}
\begin{lstlisting}[language=Coq, label={lst:ltac_ex}, caption={Example of Ltac syntax}]
Theorem lt_n_S_n : forall n : nat, n < S n.
Proof.
  intro n. destruct n.
  - apply le_n.
  - apply le_n_S. apply le_n.
Qed.
\end{lstlisting}
\end{minipage}

If we check the Gallina terms generated by Ltac for this theorem using the \lstinline{Print} command,
we can see that they are the same as in \Cref{lst:gallina_ex},
but more verbose as every function is "unfolded" into its definition.

\begin{minipage}{\linewidth}
\begin{lstlisting}[language=Coq, label={lst:ltac_ex_printed}, caption={Gallina terms generated by Ltac example}]
Print lt_n_S_n.

lt_n_S_n =
  fun n : nat =>
  match n as n0 return (n0 < S n0) with
  | 0 => le_n 1
  | S n0 => (fun n1 : nat => le_n_S (S n1) (S n1) (le_n (S n1))) n0
  end
   : forall n : nat, n < S n
\end{lstlisting}
\end{minipage}

When developing proofs using Ltac, each tactic is executed or "played" one by one,
much like an interpreter. The tactics are separated by punctuation marks.
When the use of a tactic causes the proof to depend on the solving of multiple sub-proofs (called "goals"),
we can use symbols like "-", "+", and "*" to branch into these sub-proofs and solve their goals independently.
Once a goal has been solved, we can move on the next. When there are no more goals, the proof is complete.
Coq provides us with tooling that gives us the ability to see our goals and the proof state
to further simplify the process \cite{cide}.
Ltac is not the only proof language, with another example being \emph{SSReflect} \cite{cssr}.

\subsection{Other proof assistants}
\label{ssec:other_proof_assistants}

Coq is a widely used proof assistant, but it is not the only one.
There are also other related tools that can be used for formalization and verification.

\subsubsection{Agda}
\label{sssec:agda}

Agda is a dependently typed functional programming language based on Martin-Löf's
intuitionistic type theory. Unlike Coq, Agda does not use tactics.
However, by using proposition as types, Agda can also function as a proof assistant
\cite{agdatut}.

\subsubsection{Lean}
\label{sssec:lean}

Lean is dependently typed functional programming language and theorem prover.
The goal of Lean is to establish a connection between the automated and interactive processes of proving theorems.
Thus, Lean functions both as a proof assistant and as an automated theorem prover.
Like Coq, Lean is also based on the type theory of CIC
\cite{lean}.

\subsection{Extraction of programs from verified proofs}
\label{ssec:extraction_of_programs_from_verified_proofs}

By the notion of propositions as types, we can use a proof assistant to
prove the correctness of a program \cite[Preface]{bpierce}. However, we can also extract a program from
a proof of its correctness. This type of program extraction is a common feature of proof assistants.
The extracted program is guaranteed to be correct by the type system of the proof assistant,
and the code for the program can be given as source code
in several programming languages, such as Haskell and OCaml (as is the case for Coq) \cite{cintro}.
