\chapter{Related \& Future Work}

\section{Related work: universe consistency checking in Coq}

There is an ongoing effort \cite{coqgithub} to implement a version of the algorithm from \cite{mbezem}
for use in universe consistency checking in Coq.
This version of the algorithm is graph-based, does not include joins,
and only support constraints (clauses) of the form $x + k >= y$ with $k \in \{0, 1\}$.
The main focus is on speed, which explains the limited expressivity.
There are plans to extend the current implementation with more expressivity later on.

\section{Future work}

\subsection{Completing proof of remaining logical lemmas}

As mentioned in \Cref{ssec:incomplete_formal_proofs_for_some_purely_logical_lemmas}, we have not fully completed
some of the intermediate lemmas used directly in the proof of \Cref{thm:3.2}.
We have done some work on making sure these are purely logical steps,
and that they should have no effect on the final result.
Solving these fully would however strengthen the argument for the correctness of the formal,
and hence informal, proof of \Cref{thm:3.2}.

\subsection{Formal proof of \Cref{lem:3.3}}

As mentioned in \Cref{ssec:leaving_out_the_formal_proof_of_lem_33}, we have not given a full formal proof
of \Cref{lem:3.3}. As we also saw in \Cref{sec:limitations}, constructing a formal proof of this lemma
would fix some cases where the extracted algorithm fails with an exception.
Further work on a complete formalization of \Cref{thm:3.2} would certainly benefit from a formal proof of \Cref{lem:3.3}.

\subsection{Proving minimality of model generated by \Cref{thm:3.2}}

As mentioned in \Cref{ssec:proof_of_minimality}, our proof of \Cref{thm:3.2} does not include a claim that the model
generated is minimal. As later explained in the same subsection, we would only need to include one additional
assumption in our proof of \Cref{thm:3.2} to be able to correctly claim that the model generated is minimal.
We chose not to do this as it had no effect on the results produced by the extracted algorithm,
only the claim for correctness.
