\pagenumbering{roman}

\begin{abstract}

    Proofs are an important part of mathematics, but they are not without their flaws.
    Most proofs are written by humans, and humans make mistakes.
    In this thesis, we explore the use of proof assistants to construct formal versions of informal proofs,
    and to extract certified and correct programs from these formal proofs.
    We study a specific case of two problems from lattice theory,
    solved by \citeauthor{mbezem} in \cite{mbezem}.
    Firstly, we ask ourselves: are the results from \cite{mbezem} correct?
    The informal proofs posed in \cite{mbezem}, used to justify the correctness of the results,
    are complex enough that mistakes are possible.
    Using the Coq proof assistant, we formalize parts of the proofs from \cite{mbezem},
    to gain more confidence in the correctness of these informal proofs.
    Secondly, is the process of formalization a feasible one?
    Transforming an informal proof into a formal proof is not necessarily an easy or straightforward process,
    and there are several approaches to formalization of a proof.
    Lastly, is the process of formalization worth the effort?
    A fully formalized proof gives us almost complete confidence in the correctness of the proof,
    and hence the result.
    Moreover, if the proof is constructive, we get the added bonus of extracting a certified program from the proof.
    In our case, this program has some practical use cases in a real-world setting,
    mostly as a prototype.

\end{abstract}

\renewcommand{\abstractname}{Acknowledgments}
\begin{abstract}

    First and foremost, I would like to extend a massive thank you to my supervisor,
    Marc Bezem. Without your excellent guidance, I would not have been able to
    complete this project. Thank you for all the help, and for all the knowledge you have
    shared with me.
    I would also like to thank all my friends at Jafu for always aiding in procrastination
    and for all the fun times.
    Lastly, I would like to thank my family for their support and encouragement.

	\vspace{1cm}
	\hspace*{\fill}\texttt{Andreas Salhus Bakseter}\\
	\hspace*{\fill}\today
\end{abstract}
\setcounter{page}{1}
\newpage
